Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-10 pivot/paper.tex
Properties of pivot

\item We formalize the language (Section~\ref{sec:foo}) and the type provider for data exploration
  (Section~\ref{sec:pivot}) and show that queries constructed using the type provider are valid
  (Section~\ref{sec:proofs}). Our formalization also covers the laziness of type providers, which
  is an important aspect not covered in the existing literature.
\end{itemize}

@@

\noindent
During type-checking, the type system keeps track of a list of provided class definitions $L$. 
Checking member access is then just a matter of finding the corresponding class definition and
finding the member type:
%
\begin{equation*}
\inference[(\emph{member})~]
  {L; \Gamma \vdash e : C & \kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\end{equation*}

@@

\noindent
In the following two sections, we focus on formalizing the pivot type provider and the nominally 
typed host language. We define the \ident{fields} predicate in Section~\ref{sec:proofs} and use it 
to prove properties of the pivot type provider.

@@

\begin{figure}
\begin{equation*}
\begin{array}{rcl}
  v &=& C(v)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(v)\; \lsep n \lsep s \lsep D \\
  e &=& C(e)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(e)\;\, \lsep x \lsep v \lsep e.N \lsep \ldots\\
  E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} (E) \\
    \\
 \tau &=& C \lsep \kvd{num} \lsep \kvd{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}
\vspace{-1em}

\[
\inference[(\emph{member})~]
{ L(C) = (\kvd{type}~C(x:\tau)= \ldots~\kvd{member}~N_i : \tau_i = e_i~\ldots), L'}
{ (C(v)).N_i \leadsto_L e_i[x \leftarrow v] }\\
\]
\vspace{-2.5em}

\[
\inference[(\emph{context})~]
{e \leadsto_L e'}
{E[e] \leadsto_L E[e']}
\]

\caption{Syntax and remaining reduction rules of the Foo calculus}
\label{fig:foo-syntax}
\end{figure}

@@

The Foo calculus includes two special types. \ident{Query} is a type of data and queries 
constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
models a data series mapping keys of type $\tau_1$ to values of type $\tau_2$. A series is created 
as a result of a query written using the pivot type provider. In the actual implementation, a series 
can be passed to a charting library. A series is a statically-typed wrapper over a \ident{Query} 
value and the proofs in Section~\ref{sec:proofs} show that a series obtained from the pivot type 
provider contains keys and values of matching types.

\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is prameterized by a function
$L$ that maps class names to class definitions, together with nested classes associated with the
class definition (used during type-checking as discussed below). The map is not used in the 
reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
omitted there.

@@

\subparagraph{Type checking.} One interesting aspect of type-checking with type providers is that
type providers can provide potentially infinite number of types. In F\#, the types are provided 
lazily as the type checker explores parts of the type space used by the program [X]. Consider:
%
\begin{equation*}
\ident{olympics}.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}
\end{equation*}

@@

\noindent
The judgement states that given class definitions $L_1$ and a variable context $\Gamma$, the type
of expression $e$ is $\tau$ and the type checking evaluated further class definitions that are
now included in $L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
$C$ to a pair consisting of the definition and a function that provides definitions of 
further classes: 
%
\begin{equation*}
L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L'
\end{equation*}

\noindent
The types of members of the class $C$ may be already known classes defined by $L_1$, but also classes 
defined by $L'$. This models laziness as $L'$ is a function that may never be evaluated.

The rules that define type-checking are shown in Figure~\ref{fig:foo-typecheck}. The two rules
that force the discovery of new classes are (\emph{new}) and (\emph{member}). In both cases,
we perform a lookup using $L_2(C)$ which returns a class definition together with further classes
$L$. We treat functions as sets and join $L_2$ with further classes defined by $L$ using $L_2 \cup L$.

@@

\begin{figure}[t]

\[
\inference[(\emph{num})~]{}{L; \Gamma \vdash n : \kvd{num}; L }
\quad
\inference[(\emph{string})~]{}{L; \Gamma \vdash s : \kvd{string}; L}
\quad
\inference[(\emph{var})~]{}{L; \Gamma, x:\tau\vdash x:\tau; L}
\]
\vspace{-2em}

@@

\[
\inference[(\emph{new})~]
  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots), L}
  {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
\]
\vspace{-2em}

\[
\inference[(\emph{member})~]
  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L}
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
\]
\vspace{-2em}

\[
\inference[(\emph{class})~]
  { L; \Gamma,x:\tau \vdash e_i : \tau_i; L_{i} & m_i = \kvd{member}~N_i:\tau_i= e_i }
  { L \vdash \kvd{type}~C(x:\tau) = m_1\ldots m_n; L_1 \cup \ldots \cup L_n }
\]
\caption{Type-checking of Foo expressions with lazy context}
\label{fig:foo-typecheck}
\end{figure}

@@

The rules for primitive types and variables are standard.
Input data (\emph{data}) is of type \ident{Query} and all the operations of relational algebra 
take a value of type \ident{Query} and produce a result of type \ident{Query}. This is the 
untyped (or one-typed) subset of the Foo calculus. A \ident{Query} value can be converted into a
series type (\emph{series}) of any type\footnote{This can be seen as a boundary between static and
dynamic typing in gradually typed languages [X,Y,Z].} -- in the pivot type provider, the 
\ident{Query} type is always hidden and users can only work with type-safe series type.

Finally, we also define a judgement $L_1 \vdash l; L_2$ which states that given known classes $L_1$,
a class definition $l$ is well typed and produces known classes $L_2$. This is defined by the
(\emph{class}) rule that checks the types of bodies of all members of the class.

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{pivot}(F) = C, \{ C \mapsto (l, L_1 \cup \ldots \cup L_4) \}\qquad\ball{1}\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
\qquad \qquad \kvd{member}~\qident{sort data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{sort}(F)\\
\qquad \qquad \kvd{member}~\qident{group data} : C_3 = C_3(x) &\textnormal{where}~C_3, L_3 = \ident{group}(F)\\
\qquad \qquad \kvd{member}~\qident{get series} : C_4 = C_4(x) &\textnormal{where}~C_4, L_4 = \ident{get-key}(F)\\
\\
\ident{get-key}(F) = C, \{ C \mapsto (l, \bigcup L_f)\}\qquad\ball{2} \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{with key $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{get-val}(F, f)\\
\\
\ident{get-val}(F, f_k) = \{ C \mapsto (l, \{\}) \}\qquad\ball{3} \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f_k\}~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{and value $f$} : \ident{series}\langl\tau_k, \tau_v\rangl = &\quad  \tau_k = F(f_k),~ \tau_v = F(f) \\
\qquad \qquad \quad \ident{series}\langl\tau_k, \tau_v\rangl(\Pi_{f_k, f}(x))\qquad\ball{4}
\end{array}
\end{equation*}
\vspace{-1em}

\caption{Pivot type provider -- providing entry-point type and accessing transformed data}
\label{fig:tp-main}
\end{figure}

@@

\section{Pivot type provider}
\label{sec:pivot}

@@

\subparagraph{Entry-point and data access.} Figure~\ref{fig:tp-main} shows three of the functions
defining the pivot type provider. The \ident{pivot} function \ball{1} defines the entry-point type, 
which lets the user choose which operation to perform before specifying parameters of the operation. 
This is the type of \ident{olympics} in the examples throughout this paper. The definition generates 
a new class $C$ with members that wrap the input data in further classes generated by other parts
of the type provider. The result of \ident{pivot} is the class name $C$ together with definition of 
the class and further generated types. The definition is a function that only needs to be evaluated 
when a program accesses a member of the class $C$, modelling the laziness of the type provider.
In the implementation, we return the name $C$, but only need to compute the definition when the 
type checker needs to access the body.

@@

\noindent
The \ident{get-key} function generates a class with one member for each field in the data set.
The returned class $C_f$ is generated by \ident{get-val} and lets the user choose any of the 
remaining fields as the value. The key and value columns are then selected using $\Pi_{f_k,f}$ \ball{4}.
The series is then created with a data set containing only the key and value columns (we assume 
the order of columns is preserved). Creating a series does not statically enforce that the 
data set has the right structure, but the properties discussed in Section~\ref{sec:proofs} show 
that series obtained from the pivot type provider is constructed correctly.

@@

\begin{figure}
\vspace{-0.75em}
\begin{equation*}
\begin{array}{ll}
\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
\\
\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\quad\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
\qquad  \textnormal{where}\\[0.5em]
\multicolumn{2}{l}{
~\,\qquad \begin{array}{l}
 C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
 C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
 C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
 C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
 C'', L'' = \ident{pivot}(G)
\end{array} 
}
\end{array}
\end{equation*}
\vspace{-1em}

@@

\subsection{Filtering}
\label{sec:pivot-filter}
\newpage

>>>>

\item We formalize the language (Section~\ref{sec:foo}) and the type provider for data exploration
  (Section~\ref{sec:pivot}) and show that queries constructed using the type provider are valid
  (Section~\ref{sec:pivot-prop}). Our formalization also covers the laziness of type providers, which
  is an important aspect not covered in the existing literature.
\end{itemize}

@@

\noindent
During type checking, the type system keeps track of a list of provided class definitions $L$. 
Checking member access is then just a matter of finding the corresponding class definition and
finding the member type:
%
\begin{equation*}
\inference[(\emph{member})~]
  {L; \Gamma \vdash e : C & \kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\end{equation*}

@@

\noindent
In the following two sections, we focus on formalizing the pivot type provider and the nominally 
typed host language. We define the \ident{fields} predicate in Section~\ref{sec:pivot-prop} and use it 
to prove properties of the pivot type provider.

@@

\begin{figure}
\vspace{-1em}
\begin{equation*}
\begin{array}{rcl}
  v &=& C(v)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(v)\; \lsep n \lsep s \lsep D \\
  e &=& C(e)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(e)\;\, \lsep x \lsep v \lsep e.N \lsep \ldots\\
  E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (E) \\
    \\
 \tau &=& C \lsep \kvd{num} \lsep \kvd{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}
\vspace{-1.5em}

\[
\inference[(\emph{member})~]
{ L(C) = (\kvd{type}~C(x:\tau)= \ldots~\kvd{member}~N_i : \tau_i = e_i~\ldots), L'}
{ (C(v)).N_i \leadsto_L e_i[x \leftarrow v] }\\
\]
\vspace{-3em}

\[
\inference[(\emph{context})~]
{e \leadsto_L e'}
{E[e] \leadsto_L E[e']}
\]
\caption{Syntax and remaining reduction rules of the Foo calculus}
\label{fig:foo-syntax}
\end{figure}

@@

The Foo calculus includes two special types. \ident{Query} is a type of data and queries 
constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
models a data series mapping keys of type $\tau_1$ to values of type $\tau_2$. A series is created 
as a result of a query written using the pivot type provider. In the actual implementation, a series 
can be passed to a charting library. A series is a statically-typed wrapper over a \ident{Query} 
value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained from the pivot type 
provider contains keys and values of matching~types.

\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is prameterized by a function
$L$ that maps class names to class definitions, together with nested classes associated with the
class definition (used during type checking as discussed below). The map is not used in the 
reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
omitted there.

@@

\subparagraph{Type checking.} One interesting aspect of type checking with type providers is that
type providers can provide potentially infinite number of types. In F\#, the types are provided 
lazily as the type checker explores parts of the type space used by the program [X]. Consider:
%
\begin{equation*}
\ident{olympics}.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}
\end{equation*}

@@

\noindent
The judgement states that given class definitions $L_1$ and a variable context $\Gamma$, the type
of expression $e$ is $\tau$ and the type checking evaluated class definitions that are
now included in~$L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
$C$ to a pair consisting of the definition and a function that provides definitions of 
delayed classes: 
%
\begin{equation*}
L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L
\end{equation*}

\noindent
The class $C$ may use classes defined in $L_1$, but also delayed classes from $L$. This models 
laziness as $L$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
cannot check class definitions upfront as in typical object calculi [X]. Instead, we check that 
that members are well typed as they appear in the source code, which matches the behaviour of F\# 
type providers. In general, this means that $L$ may contain classes with incorrectly typed 
members. We prove that this is not the case for the pivot type provider (Section~\ref{sec:pivot-prop}).

The rules that define type checking are shown in Figure~\ref{fig:foo-typecheck}. The two rules
that force the discovery of new classes are (\emph{new}) and (\emph{member}). In (\emph{new}), we
find the class definition and delayed classes using $L_2(C)$. We treat functions as sets and 
join $L_2$ with delayed classes defined by $L$ using $L_2 \cup L$. In (\emph{member}), we obtain
the class definition and discover delayed classes in the same way, but we also check that the body
of the member is well-typed.

@@

\begin{figure}[t]
\vspace{-1em}
\[
\inference[(\emph{num})~]{}{L; \Gamma \vdash n : \kvd{num}; L }
\quad
\inference[(\emph{string})~]{}{L; \Gamma \vdash s : \kvd{string}; L}
\quad
\inference[(\emph{var})~]{}{L; \Gamma, x:\tau\vdash x:\tau; L}
\]
\vspace{-2em}

@@

\[
\inference[(\emph{new})~]
  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots),L}
  {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
\]
\vspace{-2em}

\[
\inference[(\emph{member})~]
  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2\cup L; \Gamma,x:\tau \vdash e_i : \tau_i; L_3 \\
   L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L }
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_3}
\]
\vspace{-1em}
\caption{Type-checking of Foo expressions with lazy context}
\label{fig:foo-typecheck}
\end{figure}

@@

The rules for primitive types and variables are standard. Input data (\emph{data}) is of type 
\ident{Query} and all the operations of relational algebra take a value of type \ident{Query} and 
produce a result of type \ident{Query}. An untyped \ident{Query} value can be converted into a 
series type (\emph{series}) of any type -- in the pivot type provider, the \ident{Query} type is 
always hidden and users can only work with type-safe series type. The bridging of typed and 
untyped parts of the language resembles the boundary between static and dynamic typing in 
gradually typed languages.% [X,Y,Z].

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{pivot}(F) = C, \{ C \mapsto (l, L_1 \cup \ldots \cup L_4) \}\qquad\ball{1}\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
\qquad \qquad \kvd{member}~\qident{sort data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{sort}(F)\\
\qquad \qquad \kvd{member}~\qident{group data} : C_3 = C_3(x) &\textnormal{where}~C_3, L_3 = \ident{group}(F)\\
\qquad \qquad \kvd{member}~\qident{get series} : C_4 = C_4(x) &\textnormal{where}~C_4, L_4 = \ident{get-key}(F)\\
\\
\ident{get-key}(F) = C, \{ C \mapsto (l, \bigcup L_f)\}\qquad\ball{2} \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{with key $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{get-val}(F, f)\\
\\
\ident{get-val}(F, f_k) = C, \{ C \mapsto (l, \{\}) \}\qquad\ball{3} \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f_k\}~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{and value $f$} : \ident{series}\langl\tau_k, \tau_v\rangl = &\quad  \tau_k = F(f_k),~ \tau_v = F(f) \\
\qquad \qquad \quad \ident{series}\langl\tau_k, \tau_v\rangl(\Pi_{f_k, f}(x))\qquad\ball{4}
\end{array}
\end{equation*}

\caption{Pivot type provider -- entry-point type and accessing transformed data}
\label{fig:tp-main}
\end{figure}

@@

\section{Formalising the pivot type provider}
\label{sec:pivot}

@@

\subparagraph{Entry-point and data access.} Figure~\ref{fig:tp-main} shows three of the functions
defining the pivot type provider. The \ident{pivot} function \ball{1} defines the entry-point type, 
which lets the user choose which operation to perform before specifying parameters of the operation. 
This is the type of \ident{olympics} in the examples throughout this paper. The definition generates 
a new class $C$ with members that wrap the input data in delayed classes generated by other parts
of the type provider. The result of \ident{pivot} is the class name $C$ together with definition of 
the class and delayed generated types. The definition is a function that only needs to be evaluated 
when a program accesses a member of the class $C$, modelling the laziness of the type provider.
In the implementation, we return the name $C$, but only need to compute the definition when the 
type checker needs to access the body.

@@

\noindent
The \ident{get-key} function generates a class with one member for each field in the data set.
The returned class $C_f$ is generated by \ident{get-val} and lets the user choose any of the 
remaining fields as the value. The key and value columns are then selected using $\Pi_{f_k,f}$ \ball{4}.
The series is then created with a data set containing only the key and value columns (we assume 
the order of columns is preserved). Creating a series does not statically enforce that the 
data set has the right structure, but the properties discussed in Section~\ref{sec:pivot-prop} show 
that series obtained from the pivot type provider is constructed correctly.

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
\\
\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
\qquad  \textnormal{where}\\[0.5em]
\multicolumn{2}{l}{
~\,\qquad \begin{array}{l}
 C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
 C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
 C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
 C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
 C'', L'' = \ident{pivot}(G)
\end{array} 
}
\end{array}
\end{equation*}

@@

\newpage

\subsection{Properties of the pivot type provider}
\label{sec:pivot-prop}

If we were using the relational algebra formalised in Section~\ref{sec:foo-rel} to construct 
queries, we can obtain an invalid program, e.g.~by attempting to select
a column $f$ using $\Pi_{f}$ from a data set that does not contain the column. This is not an issue 
when using the pivot type provider, because the provided types allow the user to construct only 
correct data transformations.

To formalise this, we prove partial soundness of the Foo calculus (Theorem~\ref{thm:foo-sound}),
which characterises the invalid programs that can be written using the \ident{Query}-typed 
expressions and then prove safety of the pivot type provider (Theorem~\ref{thm:pivot-safe}),
which shows that such errors do not occur when using the provided types.

\subparagraph{Foo calculus.} The Foo calculus consists of the relational algebra and simple object
calculus where objects can be constructed and their members accessed. It permits recursion as a
member can invoke itself on a new object instance. To accommodate this, we formalise soundness 
using progress (Lemma~\ref{thm:foo-progress}) and preservation (Lemma~\ref{thm:foo-pres}).

The soundness is partial because the evaluation can get stuck when an operation of the relational 
algebra on a given data set is undefined.

\begin{theorem}[Partial soundness]
\label{thm:foo-sound}
For all $L_0, e, e'$, if $L_0, \emptyset \vdash e : \tau, L_1$ and $e\leadsto_{L_1}^{*} e'$ 
then either $e'$ is a value, or there exists $e''$ such that $e'\leadsto_{L_1}^{*} e''$, or
$e'$ has one of the following forms:
$E[\Pi_{f_1, \ldots, f_n}(D)]$, $E[\sigma_\varphi(D)]$, $\tau_{f_1, \ldots, f_n}(D)]$ or 
$E[\Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (D)]$ for some $E, D$.
\end{theorem}
\begin{proof}
Direct consequence of Lemma~\ref{thm:foo-progress} and Lemma~\ref{thm:foo-pres}.
\end{proof}

\begin{lemma}[Partial progress]
\label{thm:foo-progress}
For all $L_0, e$ such that $L_0, \emptyset \vdash e : \tau, L_1$ then either, $e$ is a value,
there exists $e'$ such that $e\leadsto_{L_1} e'$ or $e$ has one of
the following forms: $E[\Pi_{f_1, \ldots, f_n}(D)]$, $E[\sigma_\varphi(D)]$, $\tau_{f_1, \ldots, f_n}(D)]$ 
or $E[\Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (D)]$ for some $E$ and $D$.
\end{lemma}
\begin{proof}
By induction over $\vdash$. For data, strings and numbers, the expression is always a value. 
For relational algebra operations, the expression can either be reduced or has one of the required 
forms. For (\emph{member}) the expression can always be reduced.
\end{proof}

\begin{lemma}[Type preservation]
\label{thm:foo-pres}
For all $L_0, e, e'$ such that $L_0, \emptyset \vdash e : \tau, L_1$ and $e\leadsto_{L_1} e'$ then
$L_1, \emptyset \vdash e' : \tau, L_2$ for some $L_2$.
\end{lemma}
\begin{proof}
By induction over $\leadsto_{L_1}$. Cases for relational algebra operations and for (\emph{context})
are straightforward. The (\emph{member}) case follows from a standard substitution lemma and the 
fact that type checking of member access also type checks the body of the member.
\end{proof}

\subparagraph{Correctness of the pivot provider.} 
The pivot type provider defined by \ident{pivot} defines an entry-point class and a context $L$ 
containing delayed classes. The type system cannot check type definitions in $L$ upfront, but 
we prove that the body of all provided members is well-typed. Furthermore, type checking could fail
if a delayed class was not discovered before it is needed in the (\emph{new}) and (\emph{member}) 
typing rules (Figure~\ref{fig:foo-typecheck}). We show that this cannot happen for the context
constructed by the \ident{pivot} function. To avoid operating over potentially infinite contexts,
we first define an expansion operation $\downarrow_n L$ that evaluates the first $n$ levels of the 
nested context $L$ and flattens it.

\begin{definition}[Expansion]
\label{thm:pivot-well}
Given a context $L$, we define $n^{\textnormal{th}}$ expansion of $L$, written 
$\downarrow_n L$:
\begin{itemize}
\item $\downarrow_n L = \downarrow_{n-1} L \cup \bigcup L_n$ assuming that $\downarrow_{n-1} L = \{ C_0\mapsto(l_0, L_0),\ldots,C_n\mapsto(l_n, L_n) \}$
\item $\downarrow_0 L = L$  
\end{itemize}
\end{definition}

\begin{theorem}[Correctness of lazy contexts]
\label{thm:pivot-lazy}
Given $C,L=\ident{pivot}(F)$ then for any $e$ if there exists $i, \tau$ such that 
$\downarrow_i L; \emptyset \vdash e : \tau; L'$ then also
$L; \emptyset \vdash e : \tau; L''$.
\end{theorem}
\begin{proof}
Assume there exists $F,e,i$ such that $\downarrow_i L; \emptyset \vdash e : \tau; L'$ but
not $L; \emptyset \vdash e : \tau; L''$. This is a contradiction as (\emph{new}) and (\emph{member})
typing rules expand $L$ defined by \ident{pivot} sufficiently to discover all types that may have
been used in the type-checking of $e$ using $\downarrow_i L$.
\end{proof}

\begin{theorem}[Correctness of provided types]
\label{thm:pivot-well}
For all $F,n$ let $C_0,L_0=\ident{pivot}(F)$ and assume that $C \in \dom(\downarrow_{n} L)$ 
where $\downarrow_{n} L(C) = (\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i:\tau_i= e_i\;..), L'$.
It holds that for all $i$ the body of $N_i$ is well-typed, i.e.~
$L \cup L'; x:\tau \vdash e_i : \tau_i; L''$.
\end{theorem}
\begin{proof}
By examination of the functions defining the type provider; the expressions $e_i$ are well-typed
and use only types defined in $L\cup L'$.
\end{proof}

\subparagraph{Safety of provided transformations.}
The two properties discussed above ensure that the types provided by the pivot type provider can 
be used to type check expressions constructed by the users of the type provider in the expected
way. An expression will not fail to type check because of an error in the provided types.

Now we can turn to the key theorem of the paper, which states that any expression constructed 
using (just) the provided types can be evaluated to a value of correct type. For simplicity, we
only assume expressions that access a series using the \qident{get series} member. However, this
covers all data transformations that can be constructed using the type provider.

\begin{theorem}[Safety of pivot type provider]
\label{thm:pivot-safe}
Given a schema $F=\{f_1\mapsto\tau_1, \ldots, f_n\mapsto\tau_n \}$, let $C,L=\ident{pivot}(F)$ then for any 
expression $e$ that does not contain relational algebra operations or $\ident{Query}$-typed values as sub-expression, 
if $L;x\!:\!C \vdash e : \ident{series}\langl\tau_1,\tau_2\rangl; L'$ then for all $D = 
\{ f_1\mapsto\vect{v_{1,1},\ldots,v_{1,m}}, \ldots, f_n\mapsto\vect{v_{n,1},\ldots,v_{n,m}} \}$
such that $\vdash v_{i, j} : \tau_i$ it holds that $e[x\leftarrow C(D)]\leadsto_{L'}^{*} 
  \ident{series}\langl\tau_k,\tau_v\rangl(\{ f_k\mapsto{k_1,\ldots,k_r}, f_v\mapsto{v_1,\ldots,v_r} \})$
  such that for all $j$ $\vdash k_j : \tau_k$ and $\vdash v_j : \tau_v$.
\end{theorem}
\begin{proof}
Define a mapping $\ident{fields}(C)$ that returns the fields expected in the data set passed to a 
class $C$ provided by the pivot type provider. Let $\ident{fields}(C) = F$ for $C$ provided using:
\begin{equation*}
\begin{array}{l}
\begin{array}{lcl}
\ident{pivot}(F) = C, L &&
\ident{get-key}(F) = C, L \\
\ident{drop}(F) = C, L &&
\ident{get-val}(F, f_k) = C, L\\
\end{array}
\quad\begin{array}{l}
\ident{sort}(F) = C, L\\
\ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, L\\
\end{array}\\[-0.1em]
\;\, \ident{group}(F) = C, L \;\;\,\quad \ident{agg}(F,f,G,S) = C, L
\end{array}
\end{equation*}

\noindent
By induction over $\leadsto_{L'}$, show that when $C(v).N_i$ is reduced using (\emph{member}) then 
$v$ is a value $\{ f_1\mapsto\vect{v_{1,1},\ldots,v_{1,m}}, \ldots, f_n\mapsto\vect{v_{n,1},\ldots,v_{n,m}} \}$
s.t. $\ident{fields}(C)=\{f_1\mapsto\tau_1, \ldots, f_n\mapsto\tau_n \}$ and 
$\vdash v_{i, j} : \tau_i$. Thus the class provided by \ident{get-val} is constructed with 
a data set containing the required columns of corresponding types.
\end{proof}
